Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2024-12-14 Paper/paper.tex
Changed name to iterative reduction

\title{Linear Reduction: Parsing and Interpreting Code Without Trees}
\author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
\institute{Bar-Ilan University, Ramat Gan, Israel}

@@

We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
We also create an interpreter for LLang using menhir\cite{menhir}.
In this section we detail the implementation of our algorithm and compare and contrast the two implementations.

@@

Linear reduction can be used to just parse, instead of parsing and interpreting simultaneously.
This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
Furthermore, this means linear reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.

@@

In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\cite{menhir}).
A few notes beforehand:
\begin{itemize}
    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
            on code which the menhir implementation does not support.
    \item LLang supports left composition.
            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
            We found no way of supporting this with menhir.
            This is because our grammar for the menhir implementation contains the following fragment:
            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:

@@

We begin with a simple program to demonstrate currying.
Currying is the action of taking a function $f\colon X\times Y\longto Z$ and converting it to a function $c(f)\colon X\longto(Y\longto Z)$ where $c(f)(x)\colon y\mapsto f(x,y)$.
Our LLang code is as follows:

@@

Comparing the time it takes both implementations to run this 300 times, we see that it takes linear reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.

@@

\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies (figure \ref{fig:expr1}) and
expressions generated by $\hbox{\tencsc Generate-Expression}(n,1,100)$ as $n$ varies (figure \ref{fig:expr2}).

@@

The fibonacci sequence is a linear recurrence defined by $F_n=F_{n-1}+F_{n-2}$ for $n>1$ and $F_1=F_0=1$.
We wrote a script to compute the $n$th fibonacci number recursively in LLang:

@@

We can see that linear reduction is significantly slower than the menhir-based interpreter.
This can be for various reasons:
\begin{enumerate}
    \item Linear reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
        This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
        This can be made more efficient: instead of starting the next $\beta$-reduction at the beginning of a string, it can be started at the token before the one just reduced.
        Meaning once $\s s^{n-1}$ and $\s s^n$ are reduced to give $\s x^{n-1}$, we can begin the next iteration of the $\beta$-reducer at $\s s^{n-2}\s x^{n-1}$.
        Unfortunately, implementing this did not lead to substantial increases in efficiency.

    \item When handling function calls, linear reduction simply re-reduces the string in the definition of the function.
        This requires it to parse and interpret the function again.
        But classically, since the function definition has already been parsed and a parse tree created, further parsing of the function is not created.
        This means that classical methods handle function calls better than linear reduction, but on the other hand linear reduction can handle macro-like programming better.
        For example, if one were to change the state to map \ttt( to $\Lbrace$, the menhir-based interpreter would not be able to reparse the function to reinterpret \ttt(.
        But this naturally happens with linear reduction, for better or worse.
        \gobble)\gobble)

    \item The standard methods of parsing code are well-studied and optimized.
        Menhir does not naively parse code, it is an optimized tool.
        Thus comparing a naive implementation of linear reduction to menhir is not a fair comparison.
\end{enumerate}

>>>>

\title{Iterative Reduction: Parsing and Interpreting Code Without Trees}
\author{Ari Feiglin\inst{1}\and Yoni Zohar\inst{1}}
\institute{Bar-Ilan University, Ramat Gan, Israel}

@@

We implement our algorithm in an interpreter for a toy language we dub ``ILang''.
We also create an interpreter for ILang using menhir\cite{menhir}.
In this section we detail the implementation of our algorithm and compare and contrast the two implementations.

@@

Iterative reduction can be used to just parse, instead of parsing and interpreting simultaneously.
This has its benefits, as separating parsing and interpretation allows for the inspection of code before running it, which can help alert the programmer to syntax errors.
Furthermore, this means iterative reduction can be used in the classical interpreter and compiler pipeline, as it can be used as a parser.

@@

In this section we compare the runtimes of both implementations (one using iterative reduction and one using menhir\cite{menhir}).
A few notes beforehand:
\begin{itemize}
    \item The iterative reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
            on code which the menhir implementation does not support.
    \item ILang supports left composition.
            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
            We found no way of supporting this with menhir.
            This is because our grammar for the menhir implementation contains the following fragment:
            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:

@@

We begin with a simple program to demonstrate currying.
Currying is the action of taking a function $f\colon X\times Y\longto Z$ and converting it to a function $c(f)\colon X\longto(Y\longto Z)$ where $c(f)(x)\colon y\mapsto f(x,y)$.
Our ILang code is as follows:

@@

Comparing the time it takes both implementations to run this 300 times, we see that it takes iterative reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.

@@

\noindent We then run ILang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies (figure \ref{fig:expr1}) and
expressions generated by $\hbox{\tencsc Generate-Expression}(n,1,100)$ as $n$ varies (figure \ref{fig:expr2}).

@@

The fibonacci sequence is a linear recurrence defined by $F_n=F_{n-1}+F_{n-2}$ for $n>1$ and $F_1=F_0=1$.
We wrote a script to compute the $n$th fibonacci number recursively in ILang:

@@

We can see that iterative reduction is significantly slower than the menhir-based interpreter.
This can be for various reasons:
\begin{enumerate}
    \item Iterative reduction is an $O(n^2)$-time algorithm (when no reduction produces a string of printable tokens, and every reduction reduces two tokens to one).
        This is because a string of the form $\s s^1_1\s s^2_2\cdots\s s^n_n$ will take $n$ applications of the $\beta$-reducer, each application taking $n-k$ iterations (since $\beta$ is recursive).
        This can be made more efficient: instead of starting the next $\beta$-reduction at the beginning of a string, it can be started at the token before the one just reduced.
        Meaning once $\s s^{n-1}$ and $\s s^n$ are reduced to give $\s x^{n-1}$, we can begin the next iteration of the $\beta$-reducer at $\s s^{n-2}\s x^{n-1}$.
        Unfortunately, implementing this did not lead to substantial increases in efficiency.

    \item When handling function calls, iterative reduction simply re-reduces the string in the definition of the function.
        This requires it to parse and interpret the function again.
        But classically, since the function definition has already been parsed and a parse tree created, further parsing of the function is not created.
        This means that classical methods handle function calls better than iterative reduction, but on the other hand iterative reduction can handle macro-like programming better.
        For example, if one were to change the state to map \ttt( to $\Lbrace$, the menhir-based interpreter would not be able to reparse the function to reinterpret \ttt(.
        But this naturally happens with iterative reduction, for better or worse.
        \gobble)\gobble)

    \item The standard methods of parsing code are well-studied and optimized.
        Menhir does not naively parse code, it is an optimized tool.
        Thus comparing a naive implementation of iterative reduction to menhir is not a fair comparison.
\end{enumerate}
